---
title: "Test RapidBinning"
author: "Pere Rafols"
date: "2026-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Start by setting the data_path variable pointing to the folder containing the data.
data_path <- "~/MSI_DATASETS/rMSI2_workshop"

data_desc <- rMSI2::ImzMLDataDescription() #Create an empty data description object
data_desc$setOutputPath(file.path(data_path,"out")) #Set the output data path
data_desc$fixBrokenUUID <- T #Ideally leave it to False (the default)
#Append an imzML to process. We can call this method multiple times to append more imzML file to the same processing pipeline (data merging).
data_desc$appendImzMLDataPath(file.path(data_path,"0330_neg_DAN_cerebellum_MSI_2_lowrange.imzML")) 
```



Process data using the old Binning method. Note that no peak-picking is needed!
```{r, eval=T}
proc_params <- rMSI2::ProcessingParameters() #Create a processing parameters object prefilled with the default settings.
proc_params$preprocessing$smoothing$enable <- F
proc_params$preprocessing$massCalibration <- F
proc_params$preprocessing$alignment$enable <- F
proc_params$preprocessing$peakpicking$enable <- T
proc_params$preprocessing$peakpicking$SNR <- 5
proc_params$preprocessing$peakbinning$rapid_peaks <- F
proc_params$preprocessing$peakbinning$enable <- T
proc_params$preprocessing$peakbinning$tolerance_in_ppm <- F
proc_params$preprocessing$peakbinning$tolerance <- as.integer(4) #Set the binning tolerance to avoid merging peaks from different species

start_time <- Sys.time()
rMSI2::ProcessImages( proc_params, data_desc, create_rMSIXBin_files = T ) # This will take 2 minutes
end_time <- Sys.time()
elapsed_oldBinning <- end_time - start_time
print(elapsed_oldBinning)
```

```{r, eval=T}
pkMatLegacy <- rMSI2::LoadPeakMatrix(file.path(data_path, "out", "merged-peakmatrix-legacy.pkmat"))
pkMatLegacy_ram_mb <- as.numeric(object.size(pkMatLegacy$intensity)) / (1024^2)
cat(sprintf("Memory used by leagacy peak matrix: %.2f MB\n", pkMatLegacy_ram_mb))
```

Process data using the new Rapid Binning method. Note that no peak-picking is needed!
```{r, eval=T}
proc_params <- rMSI2::ProcessingParameters() #Create a processing parameters object prefilled with the default settings.
proc_params$preprocessing$smoothing$enable <- F
proc_params$preprocessing$massCalibration <- F
proc_params$preprocessing$alignment$enable <- F
proc_params$preprocessing$peakpicking$enable <- F #No need to use the full image peak picking with the new binning approach
proc_params$preprocessing$peakbinning$rapid_peaks <- T #This is the key to use the new binning approach
proc_params$preprocessing$peakbinning$max_rapidPeaksMEM_MB <- pkMatLegacy_ram_mb #Max memory allocated to store the peak matrix
proc_params$preprocessing$peakbinning$tolerance_in_ppm <- F
proc_params$preprocessing$peakbinning$tolerance <- as.integer(4) #Set the binning tolerance to avoid merging peaks from different species

start_time <- Sys.time()
rMSI2::ProcessImages( proc_params, data_desc ) # This will take 2 minutes
end_time <- Sys.time()
elapsed_rapidBinning <- end_time - start_time
print(elapsed_rapidBinning)


pkMatRapid <- rMSI2::LoadPeakMatrix(file.path(data_path, "out", "merged-peakmatrix.pkmat"))
```


Time diffs:
```{r, eval=T}
cat(sprintf("Time diff old - rapid = %s\n", format(elapsed_oldBinning - elapsed_rapidBinning)))
cat(sprintf("Performance gain: %.2f%% time reduction\n", 
            ((as.numeric(elapsed_oldBinning, units = "secs") - as.numeric(elapsed_rapidBinning, units = "secs")) / as.numeric(elapsed_oldBinning, units = "secs")) * 100))

```


DRAFT run of the new Cpp implementation
```{r, eval=F}
#Load MSI data
start_time <- Sys.time()
img <- rMSI2::LoadMsiData("~/MSI_DATASETS/rMSI2_workshop/0330_neg_DAN_cerebellum_MSI_2_lowrange.imzML")
#img <- rMSI2::LoadMsiData("~/MSI_DATASETS/cbookmeyer/2022-Q2 test MALDI/0420_cerebellum_neg_DAN_LMWCs.imzML")
end_time <- Sys.time()
elapsed_loadMSI <- end_time - start_time
print(elapsed_loadMSI)


#Run the new binning C++ backend
start_time <- Sys.time()
pksC <- rMSI2::FastPeakBinning(img_lst = list(img), min_SNR = 0.5, max_mem_MB =  27.3802, peak_width_scans = 4, n_cores = parallel::detectCores() )
end_time <- Sys.time()
elapsed_rapidBinning_C <- end_time - start_time
print(elapsed_rapidBinning_C)

#Compare R vs. C matrices
pkMatRapid <- pksC

identical(pkMatLegacy$intensity, pkMatRapid$intensity)

```



Compare information in old vs. new peak matrices

Single ion image test
```{r, eval=T}
test_ion_mz <- 281

#Old peak mat
rMSI2::plotPeakImage(pkMatLegacy, mz = test_ion_mz, labels = "Legacy" )

#New peak mat without peak-picking
rMSI2::plotPeakImage(pkMatRapid, mz =  test_ion_mz, labels = "Rapid" )

```


Compare both matrices using PCA:
```{r, eval=T}
pca_old <- prcomp(pkMatLegacy$intensity, scale. = F, center = F)
pca_new <- prcomp(pkMatRapid$intensity, scale. = F, center = F)


library(ggplot2)
dfPCA <- data.frame(PC1 = c(pca_old$x[,1], pca_new$x[,1]), 
                    PC2 = c(pca_old$x[,2], pca_new$x[,2]), 
                    group = c(rep("old", nrow(pca_old$x)), rep("new", nrow(pca_new$x))) )
dfPCA <- dfPCA[sample(1:nrow(dfPCA)), ]
scorePlot <- ggplot(dfPCA, aes(PC1, PC2))  + theme_bw()
scorePlot <- scorePlot + geom_point(aes(colour = factor(group)), alpha = 0.2, size = 1)
#scorePlot <- scorePlot + scale_colour_manual(values = clusterColors, name = "Cluster")
scorePlot <- scorePlot + guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.8)))
scorePlot

for(pci in 1:5)
{
  rMSI2::plotValuesImage(pkMatLegacy, pca_old$x[,pci], labels = paste0("PC", pci, " Legacy") ) 
  rMSI2::plotValuesImage(pkMatRapid, pca_new$x[,pci], labels = paste0("PC", pci, " Rapid") )   
}
```

Compare peak matrices using UMAP projections
```{r, eval=T}
library(uwot)

# Run fast multi-threaded UMAP (returns matrix directly)
umap_old_layout <- uwot::umap(pkMatLegacy$intensity, n_components = 5, n_threads = parallel::detectCores() - 1)
umap_new_layout <- uwot::umap(pkMatRapid$intensity,  n_components = 5, n_threads = parallel::detectCores() - 1)

# Plot spatial images
for (pci in 1:5)
{
  rMSI2::plotValuesImage(pkMatLegacy, umap_old_layout[, pci], labels = paste0("UMAP", pci, " Legacy"))
  rMSI2::plotValuesImage(pkMatRapid, umap_new_layout[, pci], labels = paste0("UMAP", pci, " Rapid"))
}
```

Check mass accuracy for the feat_list agains the old 
```{r, eval=T}
feat_list <- data.frame( mass = pkMatRapid$mass, mass_old = NA, err_ppm = NA)
pksmass_cpy <- pkMatLegacy$mass
for( i in 1:nrow(feat_list))
{
  id <- which.min(abs(pksmass_cpy - feat_list$mass[i] ))
  
  cur_err <- 1e6 * (pksmass_cpy[id] - feat_list$mass[i]) / pksmass_cpy[id]
  
  dupId <- which(feat_list$mass_old == pksmass_cpy[id])
  if(length(dupId) > 0 )
  {
    if(abs(cur_err) < abs(feat_list$err_ppm[dupId]))
    {
      
      feat_list$mass_old[dupId] <- NA
      feat_list$err_ppm[i] <- cur_err
      feat_list$err_ppm[dupId] <- NA
    }
  }
  else
  {
    feat_list$mass_old[i] <- pksmass_cpy[id]
    feat_list$err_ppm[i] <- cur_err
  }

}

target_max_error <- 2
nokids <- which(abs(feat_list$err_ppm) > target_max_error)
plot(feat_list$mass_old, feat_list$err_ppm, col = "red", pch = 20, xlab = "m/z", ylab = "error [ppm]", main = "m/z Error Legacy vs. Rapid ")
points(feat_list$mass_old[-nokids], feat_list$err_ppm[-nokids], col = "blue", pch = 20)
abline(h = c(-target_max_error,target_max_error), col = "gray")


####nokids_featList <- 
for(nokid in nokids)
{
  #Old peak mat
  rMSI2::plotPeakImage(pkMatLegacy, mz = feat_list$mass_old[nokid], labels = "Legacy" )
  
  #New peak mat without peak-picking
  rMSI2::plotPeakImage(pkMatRapid, mz =  feat_list$mass[nokid], labels = "Rapid" )
}

```
Launch MSI windows to see what happens at peak m/z 328.1351
```{r, eval=F}

#Load RAW DATA
img <- rMSI2::LoadMsiData(file.path(data_path,"0330_neg_DAN_cerebellum_MSI_2_lowrange.imzML"))
#rMSI2::MSIWindow(img)

topTIC_rows <- order(img$normalizations$TIC, decreasing = TRUE)[1:20]
spc <- rMSI2::loadImgChunkFromIds(img, topTIC_rows)

library(ggplot2)
library(reshape2)

# Convert matrix to long format
df_spc <- as.data.frame(t(spc))
df_spc$mass <- img$mass

df_long <- melt(df_spc, id.vars = "mass", variable.name = "spectrum", value.name = "intensity")

# Define target m/z window
mz_OldPeak <- feat_list$mass_old[nokid[1]]
mz_NewPeak <- feat_list$mass[nokid[1]] 
target_mz <- feat_list$mass[nokid[1]]
mz_window <- 0.02
mz_min    <- target_mz - mz_window
mz_max    <- target_mz + mz_window

# Filter data inside the window to compute dynamic Y limits
df_window <- df_long[df_long$mass >= mz_min & df_long$mass <= mz_max, ]

y_min <- 0  # Set to min(df_window$intensity) if negative values exist
y_max <- max(df_window$intensity, na.rm = TRUE) * 1.05 # Add 5% padding for top space

# Plot centered on window with auto-scaled Y-axis
ggplot(df_long, aes(x = mass, y = intensity, color = spectrum)) +
  geom_line(alpha = 0.7) +
  # Add vertical reference lines
  geom_vline(xintercept = mz_OldPeak, color = "red", linetype = "dashed", size = 0.6) +
  geom_vline(xintercept = mz_NewPeak, color = "blue", linetype = "dashed", size = 0.6) +
  # Optional text labels above the lines
  annotate("text", x = mz_OldPeak, y = y_min, label = sprintf("Old %.4f", mz_OldPeak), 
           color = "red", angle = 90, vjust = -0.5, hjust = 0, size = 3) +
  annotate("text", x = mz_NewPeak, y = y_min, label = sprintf("New %.4f", mz_NewPeak), 
           color = "blue", angle = 90, vjust = -0.5, hjust = 0, size = 3) +
  coord_cartesian(
    xlim = c(mz_min, mz_max),
    ylim = c(y_min, y_max)
  ) +
  theme_bw() +
  labs(
    x = "m/z", 
    y = "Intensity", 
    title = sprintf("Spectra centered at m/z %.4f (±0.5 Da)", target_mz)
  ) +
  theme(legend.position = "none")

```

TODO:
-> Modify the GUI with the new settings
-> create drifted mass test-data (using same data) to check merged and non-merged modes
-> Rethink about mass calibration: at the end, with other reference spectrum...etc

