---
title: "rMSI2-Workshop"
author: "Pere Rafols"
date: '2022-06-06'
output: html_document
---

## Demo data
This R markdown script show the fundamentals of rMSI2 data visualization and processing. The demo MSI data used here can be obtained in the link below:
[RAW data imzML file](http://gofile.me/3VACP/9fCg0TRSB)
[RAW data ibd file](http://gofile.me/3VACP/AIwy9LKR3)
[Multiple tissues peak matrix](http://gofile.me/3VACP/PkWMyQpy1)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Start by setting the data_path variable pointing to the folder containing the demo data in your computer.
data_path <- "~/MSI_DATASETS/rMSI2_workshop"
```

## Inspecting MSI data

We start by loading the raw data (profile mode) to inspect it in rMSI2. To load an imzML file in rMSI2 we use the LoadMsiData() method. We recommend preceding each rMSI2's method invocation with rMSI2:: prefix to avoid calling methods with the same name from another package (like the old rMSI and rMSIproc packages).
```{r }
raw <- rMSI2::LoadMsiData(file.path(data_path,"0330_neg_DAN_cerebellum_MSI_2_lowrange.imzML")) #This will take 40 seconds
```

Next, we can launch the rMSI2 graphical user interface (GUI) to visualize the MSI data.
Note that the `eval = FALSE` parameter is used in this markdown scripts when calling GUI methods. This allows us to produce reports in html or pdf formats without calling the GUI.
```{r, eval=FALSE}
rMSI2::MSIWindow(raw)
```

Alternatively, we can use a GUI method to load the data in a user-friendly way.
```{r, eval=FALSE}
raw <- rMSI2::OpenMSI()$img1
```

## rMSIObj data structure
The following lines display the basic data structure used to handle the MSI data.
```{r, echo=FALSE}
#The name of the dataset
cat("Dataset name:\n")
raw$name

#The number of pixels in X and Y directions
cat("\nImage size:\n")
raw$size 

#The pixel size (spatial resolution) in micrometers
cat("\nPixel size:\n")
raw$pixel_size_um 

#The common mass axis to the whole dataset
cat("\nMass axis:\n")
head(raw$mass) 

#The average spectrum
cat("\nAverage spectrum:\n")
head(raw$mean) 

#The pixel coordinates for each spectrum in the dataset
cat("\nPixel coordinates:\n")
head(raw$pos)

#We can get the total number of pixels in an image using the pos matrix:
cat("\nTotal number of pixel:\n")
nrow(raw$pos)

#The intensity normalization values are stored in a data frame
cat("\nIntensity Normalization values\n")
head(raw$normalizations)

#The "data" field points to the actual spectral data and to the ion image stream. 
#This field is not designed for end-user usage. Instead, use the rMSI2 methods to access the desired data.
cat("\nThe data field\n")
summary(raw$data) 
```

## Processing MSI data

The rMSI2 processing workflow can be called using a GUI or pragmatically. Most users will probably prefer the GUI since it makes easier to set the processing parameters. 
```{r, eval=FALSE}
rMSI2::ProcessWizard() #GUI to execute the complete rMSI2 processing workflow
```

But, it may be very convenient to set the processing parameters withing an R script to automate things.
```{r, eval=FALSE}
proc_params <- rMSI2::ProcessingParameters() #Create a processing parameters object prefilled with the default settings.
proc_params$preprocessing$smoothing$kernelSize <- as.integer(5) #Set the smoothing Savitzky–Golay kernel size to 5
proc_params$preprocessing$peakpicking$SNR <- 3 #Set the SNR threshold
proc_params$preprocessing$peakpicking$WinSize <- as.integer(10) #Set the window size to roughly the peak width 
proc_params$preprocessing$peakbinning$tolerance <- as.integer(4) #Set the binning tolerance to avoid merging peaks from different species

data_desc <- rMSI2::ImzMLDataDescription() #Create an empty data description object
data_desc$setOutputPath(file.path(data_path,"out")) #Set the output data path
data_desc$appendImzMLDataPath(file.path(data_path,"0330_neg_DAN_cerebellum_MSI_2_lowrange.imzML")) #Append an imzML to process. We can call this method multiple times to append more imzML file to the same processing pipeline (data merging).

rMSI2::ProcessImages( proc_params, data_desc ) # This will take 2 minutes
```

Note that the `eval = FALSE` parameter is used when executing the workflow processing methods to prevent executing the whole processing each time we want to generate a report.

During the processing, you will be prompted by the mass calibration form to select the desired mass references. We provide an example of mass reference peaks to calibrate using DAN matrix clusters in negative ionization mode:
[DAN negative mode mass references](https://github.com/prafols/rMSI2/blob/main/massCalibrationRefs/DAN_neg.ref)

## Working with the peak matrix

After completing the data processing, we obtain a peak matrix file (.pkmat file extension). To present the peak matrix object, we will use a dataset consisting of eight mouse brain tissue sections that were combined in a single peak matrix. This dataset was acquired in a TOF spectrometer using a sputtered gold nanolayer to promote the LDI process.

First, the peak matrix object must be loaded in R's memory:
```{r}
pks <- rMSI2::LoadPeakMatrix(file.path(data_path,"MultipleBrainsPeakMatrix.pkmat"))
print(pks) #The print command shows us the peak matrix dimensions and its memory usage
```
The peak matrix follows a format similar to the MSI images data. The following code shows the basics:
```{r, echo=FALSE}
#The name of the MSI data used to construct the peak matrix. 
#Since the peak matrix can result from the merging of multiple images, we may have various names in this field.
cat("Datasets names:\n")
pks$names

#In the same sense as the names field, we have a vector with the total number of pixels contained in each image.
cat("\nDatasets number of pixels:\n")
pks$numPixels

#Similarly to the rMSIObj  the pos field contains the pixel coordinates of each observation. 
#Here, the coordinates of multiple images are combined in a single matrix.
cat("\nPixels coordinates:\n")
head(pks$pos)

#The binned peak masses common to the whole dataset.
#These masses may be used to assign each matrix column to a specific compound.
cat("\nBinned masses:\n")
head(pks$mass)

#The results of the peak picking are stored in three matrices of the same dimensions:
#- intensity: stores the peak maximum value recorded at the mass centroid position.
cat("\nIntensity matrix:\n")
head(pks$intensity[,1:4], n = 3) #Just showing the first 4 columns

#- area: stores the integrated area below the peak shape.
cat("\nArea matrix:\n")
head(pks$area[,1:4], n = 3) #Just showing the first 4 columns

#- SNR: stores the SNR detected for each peak. A given peak may show a SNR value of zero, indicating that the peak actually had an SNR value below the specified #SNR threshold, but then it was retrieved by the fill peaks algorithm.
cat("\nSNR matrix:\n")
head(pks$SNR[,1:4], n = 3) #Just showing the first 4 columns

#The intensity normalization values from the MSI data are combined into a unified data.frame.
cat("\nNormalization values:\n")
head(pks$normalizations)
```

## Basic operations using the peak matrix

First, it is very handy to directly inspect ion images from the peak matrix:
```{r, warning=FALSE}
#Ion images can be plotted, specified the approximate ion mass we are interested in:
rMSI2::plotPeakImage(pks, mz = 832)

#We can tune some plotting parameters to control the appearance
rMSI2::plotPeakImage(pks, mz = 832, rotations = c(90, 90, 90, -90, 90, 90, 90, -90), labels = paste0("img", 1:8))

#We can specify an intensity normalization for the plotting
rMSI2::plotPeakImage(pks, mz = 832, rotations = c(90, 90, 90, -90, 90, 90, 90, -90), labels = paste0("img", 1:8), normalization = "TIC")
```

Secondly, we can create peak matrices subsets using the `[ , ]` standard R operator:
```{r, warning=FALSE}
#Create a subset, containing only the first ten pixels of the peak matrix
pksSubset <- pks[1:100, ]
rMSI2::plotPeakImage(pksSubset, mz = 832, labels = "Subset 100 pixels")

#Create a subset, containing only the first ten masses of the peak matrix
pksSubset <- pks[, 1:10 ]
cat("Mass subset:\n")
pksSubset$mass

#Create a subset, containing only the first image stored in the peak matrix
pksSubset <- pks[ pks$names[1] ,  ]
rMSI2::plotPeakImage(pksSubset, mz = 832, labels = "Subset first image only")
```

Regarding intensity normalization, the rationale behind rMSI2 is not to modify data stored in the disk with normalization values and keep the data always using the raw values. This allows us to test multiple normalization approaches during the data analysis. 
To apply a normalization vector to the peak matrix, we simply divide the desired matrix by the normalization vector:
```{r, warning=FALSE}
TICnormIntensity <- pksSubset$intensity/pksSubset$normalizations$TIC
rMSI2::plotValuesImage(pksSubset, pksSubset$intensity[,1], labels = "first column RAW")
rMSI2::plotValuesImage(pksSubset, TICnormIntensity[,1], labels = "first column TIC normalized")
```

## Data analysis using the peak matrix

Going back to the dataset acquired with an Orbitrap, we are going to demonstrate how to analyze the peak matrix.
```{r}
pks <- rMSI2::LoadPeakMatrix(file.path(data_path,"out/merged-peakmatrix.pkmat"))
print(pks) #The print command shows us the peak matrix dimensions and its memory usage
```
## Discarding tissue borders
At the tissue borders, there may appear some artifacts related to OCT, matrix deposition, not enough accuracy on the acquisition region, etc.
We can use the values of the RMS normalization to select these pixels and removing them from the peak matrix.
```{r, warning=FALSE}
rMSI2::plotValuesImage(pks, pks$normalizations$RMS, labels = "RMS")

require(ggplot2)
IntensityHist <- ggplot(pks$normalizations, aes(RMS)) +
                 geom_histogram(binwidth = 1e3, colour = "blue") + scale_x_continuous(labels = function(x) format(x, scientific = TRUE)) + 
                 geom_vline(xintercept = 2.5e5, colour = "red")

IntensityHist #See the cut region

invisible( rMSI2::plotClusterImage(pks, pks$normalizations$RMS < 2.5e5, labels = "Invalid Pixels") )

#Subsetting of pks
pks <- pks[pks$normalizations$RMS > 2.5e5, ]
```

## Multivariate statistics

```{r, echo = FALSE, messages = FALSE, warning=FALSE}
#PCA example
pca <- prcomp(pks$intensity/pks$normalizations$RMS)
rMSI2::plotValuesImage(pks, pca$x[,1], labels = "PC1" ) 
rMSI2::plotValuesImage(pks, pca$x[,2], labels = "PC2" ) 
rMSI2::plotValuesImage(pks, pca$x[,3], labels = "PC3" )

#K-means example
set.seed(111) #Set a random seed to get always the same results from kmeans
clus <- kmeans(pks$intensity/pks$normalizations$RMS, centers = 5)
clusterColors <- rMSI2::plotClusterImage(pks, clus$cluster)

#PC1 vs. PC2 Score plot using same colors as in clustering plots
dfPCA <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2], cluster = clus$cluster )
scorePlot <- ggplot(dfPCA, aes(PC1, PC2))  + theme_bw()
scorePlot <- scorePlot + geom_point(aes(colour = factor(cluster)), alpha = 0.2, size = 1)
scorePlot <- scorePlot + scale_colour_manual(values = clusterColors, name = "Cluster")
scorePlot <- scorePlot + guides(color = guide_legend(override.aes = list(size = 3, alpha = 0.8)))
scorePlot

```

## Volcano plot 

```{r message = FALSE, warning = FALSE}

normMatrix <- pks$intensity/pks$normalizations$RMS
group1  <- normMatrix[clus$cluster == 3, ]  #Select observations for cluster 3
group2  <- normMatrix[clus$cluster == 4, ] #Select observations for cluster 4

#Take the same number of obervations for each group a limit them to 1000 to avoid p.values of zero value
nobs <- 1000
group1 <- group1[sample(1:nrow(group1), nobs), ]
group2 <- group2[sample(1:nrow(group2), nobs), ]

foldchange_log2 <- function(var1, var2)
{
  mv1 <- median(var1, na.rm = T)
  mv2 <- median(var2, na.rm = T)
  return (log2(mv2/mv1))
}

dfFC <- data.frame (mz = pks$mass,
                    FC =  rep(NA, length(pks$mass)),
                    pval = rep(NA, length(pks$mass)))

for( i in 1:length(pks$mass))
{
  dfFC$FC[i] <- foldchange_log2(group1[, i], group2[ , i] )
  dfFC$pval[i] <- kruskal.test(list( group1[, i], group2[ , i]))$p.value
}
dfFC$pvalAdjusted <- p.adjust(dfFC$pval, method = "BH")

#Volcano
volcanoPlot <- function( log2FoldChange, pValue, var_labels, var1_name = "var1", var2_name = "var2", threshold_fc = 0.4, threshold_p = 0.05 )
{
  require(ggplot2)
  require(ggrepel)
  vDf <- data.frame( Label = var_labels,
                     Log2.Fold.Change = log2FoldChange, 
                     P.Value = pValue, 
                     threshold = as.factor( abs(log2FoldChange) >= threshold_fc &  pValue <= threshold_p ))
  vol <- ggplot(data = vDf, aes(x=Log2.Fold.Change, y =-log10(P.Value), colour=threshold))
  vol <- vol + geom_vline( xintercept = c(-threshold_fc, threshold_fc), colour = "blue", alpha = 0.5 )
  vol <- vol + geom_hline( yintercept = -log10(threshold_p), colour = "blue", alpha = 0.5 )
  vol <- vol + geom_point(alpha=0.75, size=1.75)
  vol <- vol +  xlim(1.1*c(-max(abs(log2FoldChange)), max(abs(log2FoldChange))))
  vol <- vol + xlab( bquote("fold change {"*log[2](frac(.(var2_name), .(var1_name)) )*"}" )) + ylab(expression(-log[10]("p.value") )) + theme_bw() + theme(legend.position="none")
  vol <- vol + labs( title= paste0("Volcano ",  var2_name, " vs. ", var1_name ))
  vol <- vol + geom_text_repel( data = vDf[ which(as.logical(vDf$threshold)), ], aes( x= Log2.Fold.Change, y= -log10(P.Value), label = Label), alpha = 0.8, force = 10, size = 3, segment.size = 0.2 )
  vol
}

volcanoPlot(log2FoldChange = dfFC$FC, pValue = dfFC$pvalAdjusted, var_labels = paste0( "m/z "  , round(pks$mass, 3)), var1_name = "Cluster 3", var2_name = "Cluster 4", threshold_fc = 1  )

#Sort by Fold-Change
dfFC <- dfFC[order(dfFC$FC), ]

knitr::kable(dfFC, row.names = F, align = 'c')

```

## Images for the five most up-regulated ions
```{r warning = FALSE}
for( i in nrow(dfFC):(nrow(dfFC)-4))
{
  rMSI2::plotPeakImage(pks, mz = dfFC$mz[i], normalization = "RMS")
}
```

## Images for the five most down-regulated ions
```{r warning = FALSE}
for( i in 1:5)
{
  rMSI2::plotPeakImage(pks, mz = dfFC$mz[i], normalization = "RMS")
}
```